\textbf{Teaching a machine to generate an artificial universe}

### 2020-08-03 report.tex
formatted report.tex for github viewing

\setcounter{tocdepth}{2}
\begingroup
\let\clearpage\relax
\tableofcontents
\endgroup
\newpage

\section{Introduction}
Beyond the distance scales of galactic diameters, toward the sizes of galaxy clusters and beyond, matter in the Universe appears as a webbed structure of filaments across volumes of emptiness known as voids. This network of luminous formation is known as the cosmic web~\cite{bond_cw1, Coles_cw2, Forero_cw3, Dietrich_cw4, cw_phenomenon}. The cosmic web as a distribution of matter contains artefacts linked to the state of the primordial universe, the nature of early universe perturbations and more fundamental physics to describe this era in cosmic time. The emergence of the structure is from an early period of exponential expansion, known as inflation~\cite{inflation_cosmo2}. This structure has been observed by missions such as~\cite{sdss1, 2df} and by weak gravitational lensing~\cite{glens_book} through distortions, or shears~\cite{shear_review}, to images of distant galaxies caused by the large-scale structure of the universe.

@@

%SOME EXAMPLES OF GENERATIVE AI
Deep learning has evolved to be able to process the statistical structure of the hierarchy of features in the visual domain~\cite{nvidia_gan}. There have been a number of recent applications of generative deep learning in astrophysics. GANs have been used to create new samples of galaxy images~\cite{gal_im_gen2} and to obtain visual characteristics past the deconvolution limit~\cite{gal_im_gen3}. Other applications use the GAN framework to generate 2D images of the cosmic web~\cite{web_gan} and projected 2D mass distributions, known as convergence maps, from simulated data of weak gravitational lensing within the environment of an N-body simulation~\cite{cosmogan}. This report will show the insights that the study of astrophysics can have on AI through applications of statistics that could give new mechanisms for maximizing the realism of GAN generated samples.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% \begin{figure}
% \includegraphics[width=\columnwidth]{figures/diagrams/GAN_diagram1.png}
% \centering
% \caption{Schematic of the GAN framework; including discriminator and generator networks, the latent-space vector, dataset of samples and the backpropagation feedback to model parameters.}
% \label{fig:GAN_diagram}
% \end{figure}

@@

% WHAT ARE D AND G? HOW DO D AND G WORK?
The discriminator and generator models can be represented as functions $D$ and $G$ respectively because of their mappings as neural networks and parameter dependencies. The discriminator $D$ has input $\mathbf{x}$ and parameters $\bm{\theta}^{(D)}$. The generator $G$ takes a latent space vector $\mathbf{z}$ as input and uses parameters $\bm{\theta}^{(G)}$. Information flows through the network and its parameters to create the outputs $D(\mathbf{x})$ or $D(G(\mathbf{z}))$ for the discriminator and $G(\mathbf{z})$ for the generator. This is the forward-propagation from an input to an output. The model performances are calculated with the objective functions $J^{(D)}$ or $J^{(G)}$. %Neural networks learn through changes to their parameters and the weights are adjusted using the value of the objective function~\cite{gf_book}. 
% The process that changes the model weights to achieve a lower value of the objective function is back-propagation~\cite{backprop}. This algorithm computes the gradients of the objective function with respect to the model parameters. %Figure~\ref{fig:gan_game} shows the value function in the parameter space for which the gradients of backpropagation are calculated in the adversarial opposition of $D$ and $G$. %This means the performance of a neural network model depends on parameters $\bm{\theta}$. 
% The backpropagation adjustment to the model parameters $\bm{\theta}$ is obtained from stochastic gradient descent (SGD) (see Appendix~\ref{appendix:sgd}) of the objective function in the parameter space of either model.

@@

The mapping learnt by the generator is expressed as $G \! : \! \mathbf{z} \! \mapsto \! \mathbf{x}$. The discriminator $D$ is a map from an input of real $\mathbf{x}$ and fake samples $G(\mathbf{z})$ to an output of as many probabilities. The values of  $D(\mathbf{x})$ and $D(G(\mathbf{z}))$ signify the discriminator's confidence in the validity of the sample. This mapping is expressed as $D \! : \! \mathbf{x} \!  \mapsto \! [0,1]$. 
% The networks are differentiable with respect to their inputs and parameters because of their dependence on these objects.

@@

During training the discriminator minimises $J^{(D)}$ only through feedback from backpropagation to its parameters $\bm{\theta}^{(D)}$. The generator $G$ minimises $J^{(G)}$ by only affecting its parameters $\bm{\theta}^{(G)}$ through returns from backpropagation related to the discriminator classifications. The minimization objectives are simultaneous between two networks in search of a Nash equilibrium and added together the objectives for the discriminator and generator give a value function $V(\bm{\theta}^{(D)}, \bm{\theta}^{(G)})$ in the combined parameter space $(\bm{\theta}^{(D)}, \bm{\theta}^{(G)})$. Figure~\ref{fig:gan_game} shows the value function in the combined parameter space for which the backpropagation gradients are calculated with $D$ and $G$ as adversaries of each other. The optimum parameters for each network are shown at the Nash equilibrium. The game between the models is described with the value function $V$ as 
\begin{eqnarray}\label{eqn:value}
    \min_G \max_D V(D,G) &=& \mathbb{E}_{\mathbf{x}\sim p_{data}}\log D(\mathbf{x})                                                           \nonumber \\
                         &+& \mathbb{E}_{G(\mathbf{z}) \sim p_{model}}\log(1-D(G(\mathbf{z}))) \nonumber \\
\end{eqnarray}

@@

The Nash equilibrium is satisfied by a pair of parameter sets $(\bm{\theta}^{(D)*}, \bm{\theta}^{(G)*})$ for which no further changes occur to the objective functions $J^{(D)}$ and $J^{(G)}$ of both networks~\cite{NIPS16}. Equation~\ref{eqn:value} means that the zero-sum game for the models returns an adjustment $+V$ for the discriminator and $-V$ for the generator. Figure~\ref{fig:gan_game} shows the equilibrium point in the parameter space and the returns $V(\bm{\theta}^{(D)}, 0)$ and $V(0, \bm{\theta}^{(G)})$ for the discriminator and generator with optimum parameters $\bm{\theta}^{(D)*}$ and $\bm{\theta}^{(G)*}$ respectively in the zero-sum game. The zero-sum game between $D$ and $G$ defines an objective function $J^{(D)}$ for the discriminator
\begin{eqnarray}\label{eq:J_D}
    J^{(D)} &=& \mathbb{E}_{\mathbf{x} \sim p_{data}}\log D(\mathbf{x})  \nonumber \\
            &+&\mathbb{E}_{G(\mathbf{z}) \sim p_{model}}\log (1 - D(G(\mathbf{z})))
\end{eqnarray}

@@

\subsubsection{Experimental design of GANs} 
The opposing players in the zero-sum game are assumed by two deep convolutional neural network (DCNN) models. These networks use convolutional layers to extract features from grid-like data. Deep Convolutional Generative Adversarial Networks (DCGANs), first implemented by Radford et al. 2015~\cite{dcgan}, use these layers to create images. The DCGAN model has shown excellent results on natural image generation using the CelebA~\cite{celebA} and the LSUN-Bedrooms~\cite{LSUN} datasets. The DCNNs in this work are based on the AlexNet models for image classification~\cite{dcnn}. Despite most of the architectural features, such as convolutional layers, being used in deep learning models prior to this publication the DCGAN architecture improved previous GAN architectures for increased performance in image generation tasks. 
% See Appendix~\ref{appendix:gan_parts} for details on the relevant tools used in DCGANs. 
The models in the methods of this work borrow from the ideas of this architectural basis.

@@

\section{Applications of an artificial universe}\label{sec:universe_applications}
% WHY IS THIS IMPORTANT, WHAT COULD IT BE FOR, WHAT RESEARCH WILL COME OF IT 
\subsection{Cosmological simulations}
The Millennium Project simulation~\cite{millsim} and the successive Millennium-XXL~\cite{millxxlsim} and Millennium-II~\cite{mill2sim} simulations made foundations for the comparison of observation to theory-based simulation. Research using the simulations has provided new insight into high-redshift quiescent galaxies~\cite{mill_quiescent}, dark-matter haloes~\cite{mill_dm_haloes}, weak lensing~\cite{mill_weak_lensing}, galaxy clustering~\cite{mill_gal_clustering} and the direct comparison of galaxy surveys with galaxy groups predicted in the simulation volumes\cite{mill_sdss}. These applications in the GAN-generated galaxy distributions depend on extending the depth of the method presented in this report but they are strong motivations for the use of GANs in simulating cosmological data. The direct comparison of survey data and generated data is the more likely once the objective of this work is completed. The approach in this report can generate a similar volume to the Millennium simulation in around an hour (GPU time) as opposed to three-hundred years (CPU time) for the Millennium simulation.\footnotemark.
\footnotetext{\url{wwmpa.mpa-garching.mpg.de/galform/virgo/millennium}}

@@

The process involves the creation of a series of mock data challenges (MDCs). The MDCs are simulations of GW events in mock galaxy catalogues. Between the separate MDCs there are varying conditions of GW and EM selection effects (GW selection effects correspond to detector sensitivity and EM effects are due to observed flux limitations). The effects of clustering and large scale cosmological structure are ignored in the simulation of co-moving volumes of uniformly distributed galaxies that the GW events occur in. These mock catalogs show different comparisons to certain properties of true galaxy distributions from surveys for locating GW events, such as GLADE~\cite{GLADE}. %For the complete galaxy catalogue in one such MDC a uniform simulated distribution of galaxies is added past the limit of the catalog. 
A larger testing volume could be generated from a routine, such as the one described in this report, with the summary statistics that are similar to survey or simulation data with added realistic cosmological structure. The EM information from outside the volume specified in the catalogs could not be replaced by the GAN method shown in this report because at present it only returns positional data. This could increase the understanding of how current techniques for observed GW events perform so that tighter constraints on the source-observer separation could be obtained in the future. This would help to accumulate more data for posteriors to further constrain $H_0$.

@@

\subsection{Extraction of training data} % TRAINING DATA MANUFACTURE AND PREPROCESSING
Position vector components for the centres of cube samples in the simulation volume were drawn from a uniform distribution along each axis of each redshift-snapshot. If the addition or subtraction of half the side length of the cube to any of the components resulted in a point outside of the simulation volume, the point was discarded. The galaxy cube samples were obtained by sorting the position vectors of galaxies in the simulation that were within the cube volume. Each point inside the cube was recorded in an array. From each simulation volume of $512 \: h^{-1} \,\textup{Mpc}$ ensembles of 64 array samples of size $80 \: h^{-1} \,\textup{Mpc}$ were selected. The galaxy positions in each array for every cube were scaled to the range $[-1,+1]$. 3D-histograms with 32 bins per axis were made on each cube across the same range for each axis. The densities in each bin cell were also scaled to this range to fit the output ranges of the GAN models. 
% After testing different sizes, a cube size of $160 \: h^{-1} \,\textup{Mpc}$ was found to balance the selection of unique samples in the simulation volume against featuring the same features in the simulation structure.

@@

where $a$ is a free parameter that depends on the sparsity of the histogram densities. The value of $a=1.25$ was found to be optimal for the $160 \:  \: h^{-1} \,\textup{Mpc}$ histograms at all redshifts $z$. This parameter remains fixed during training. The value of $a$ controls the median pixel density of the scaled image. Figure \ref{fig:scaling} shows the difference between scaled images $s(x)$ and the original images $x$. The inverse transformation $s^{-1}(x)$ can be applied to obtain the original image. 
% A logarithmic scaling was used originally, but this did not provide good results with the GAN architecture that used the hyperbolic tangent ($\tanh$) activation because the logarithm of a negative number is undefined. The models using $\tanh$ activations achieved much better results than models without these activations.

@@

% \subsubsection{2D-GAN implementation} % DESCRIBE SPECIFICS OF THIS IMPLEMENTATION
% \textit{Architecture} --- 
% The networks follow the structuring of the DCGAN models~\cite{dcgan} and are written in Keras~\cite{keras}. A full description of the layers in the 2D-GAN architecture is given in Table~\ref{table:2dgan_archi} of Appendix~\ref{appendix:gan_setups}. The generator transforms a 100-dimensional vector drawn from a Gaussian prior $\mathbf{z}\sim \mathcal{N} (0,1)$ into a $32\times 32$ single-channel image. The latent-space vector is followed by a fully connected layer that is reshaped to a stack of feature maps for the two convolutional layers. Batch-normalisation is only used after the fully connected layer. Through each layer hyperbolic tangent activations are used. See Appendices~\ref{appendix:app_acts} and~\ref{app_BN} for more detail on batch-normalisation and activation functions.

% The discriminator network has two convolutional layers and two fully connected layers. The image samples have features extracted by the first two convolutional layers and the features are flattened for the final fully connected layers. Hyperbolic tangent functions are used throughout the model and a sigmoid activation function (see Appendix~\ref{appendix:app_acts}) is used for the final single neuron layer that outputs the probability for a given sample being real.

% Both networks use the same number of filters for the corresponding layers in either model. Together the discriminator and generator models have $3.8\times 10^7$ trainable parameters. The discriminator and generator objectives of Equations~\ref{eq:J_D} and \ref{eqn:J_G} are minimized using the Adam optimiser~\cite{adam}. Each model uses a learning rate of $2\times 10^{-4}$ and momentum $\beta_1 = 0.5$. The models use a batch size of 64 histograms out of a set of 128 training histograms, where the batch size is the number of samples given to the discriminator at each iteration of training. Refer to Table~\ref{table:2dgan_params} of Appendix~\ref{appendix:gan_setups} for a complete listing of the 2D-GAN hyper-parameters used in each model. The hyper-parameters of a model are set outside of training and they shape how the training proceeds.\\

% {\setlength{\parindent}{0cm}
% \textit{Training} --- GANs are notoriously difficult to train. This means that the state of the models can generate realistic samples for a single epoch at a time. The GAN was made to record the parameters of the generator model at these epochs by using a statistical check (see Section~\ref{sec:stats_and_diags}) of the generated histograms against the training data histograms in the training loop. The loss (the objective function value) and accuracy values of the discriminator for each epoch were recorded during training.
% }

>>>>

\setcounter{tocdepth}{2}
\begingroup
\let\clearpage\relax
\tableofcontents
\endgroup
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Beyond the distance scales of galactic diameters, toward the sizes of galaxy clusters and beyond, matter in the Universe 
appears as a webbed structure of filaments across volumes of emptiness known as voids. This network of luminous formation 
is known as the cosmic web~\cite{bond_cw1, Coles_cw2, Forero_cw3, Dietrich_cw4, cw_phenomenon}. The cosmic web as a 
distribution of matter contains artefacts linked to the state of the primordial universe, the nature of early universe 
perturbations and more fundamental physics to describe this era in cosmic time. The emergence of the structure is from 
an early period of exponential expansion, known as inflation~\cite{inflation_cosmo2}. This structure has been observed by 
missions such as~\cite{sdss1, 2df} and by weak gravitational lensing~\cite{glens_book} through distortions, or 
shears~\cite{shear_review}, to images of distant galaxies caused by the large-scale structure of the universe.

@@

% SOME EXAMPLES OF GENERATIVE AI
Deep learning has evolved to be able to process the statistical structure of the hierarchy of features in the visual 
domain~\cite{nvidia_gan}. There have been a number of recent applications of generative deep learning in astrophysics. 
GANs have been used to create new samples of galaxy images~\cite{gal_im_gen2} and to obtain visual characteristics past
the deconvolution limit~\cite{gal_im_gen3}. Other applications use the GAN framework to generate 2D images of the cosmic 
web~\cite{web_gan} and projected 2D mass distributions, known as convergence maps, from simulated data of weak 
gravitational lensing within the environment of an N-body simulation~\cite{cosmogan}. This report will show the insights 
that the study of astrophysics can have on AI through applications of statistics that could give new mechanisms for
maximizing the realism of GAN generated samples.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

@@

% WHAT ARE D AND G? HOW DO D AND G WORK?
The discriminator and generator models can be represented as functions $D$ and $G$ respectively because of their 
mappings as neural networks and parameter dependencies. The discriminator $D$ has input $\mathbf{x}$ and parameters 
$\bm{\theta}^{(D)}$. The generator $G$ takes a latent space vector $\mathbf{z}$ as input and uses parameters 
$\bm{\theta}^{(G)}$. Information flows through the network and its parameters to create the outputs $D(\mathbf{x})$ or 
$D(G(\mathbf{z}))$ for the discriminator and $G(\mathbf{z})$ for the generator. This is the forward-propagation from 
an input to an output. The model performances are calculated with the objective functions $J^{(D)}$ or $J^{(G)}$.

@@

The mapping learnt by the generator is expressed as $G \! : \! \mathbf{z} \! \mapsto \! \mathbf{x}$. The 
discriminator $D$ is a map from an input of real $\mathbf{x}$ and fake samples $G(\mathbf{z})$ to an output of 
as many probabilities. The values of  $D(\mathbf{x})$ and $D(G(\mathbf{z}))$ signify the discriminator's confidence 
in the validity of the sample. This mapping is expressed as $D \! : \! \mathbf{x} \!  \mapsto \! [0,1]$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

During training the discriminator minimises $J^{(D)}$ only through feedback from backpropagation to its parameters 
$\bm{\theta}^{(D)}$. The generator $G$ minimises $J^{(G)}$ by only affecting its parameters $\bm{\theta}^{(G)}$ through 
returns from backpropagation related to the discriminator classifications. The minimization objectives are simultaneous 
between two networks in search of a Nash equilibrium and added together the objectives for the discriminator and generator 
give a value function $V(\bm{\theta}^{(D)}, \bm{\theta}^{(G)})$ in the combined parameter space $(\bm{\theta}^{(D)}, 
\bm{\theta}^{(G)})$. Figure~\ref{fig:gan_game} shows the value function in the combined parameter space for which the 
backpropagation gradients are calculated with $D$ and $G$ as adversaries of each other. The optimum parameters for each 
network are shown at the Nash equilibrium. The game between the models is described with the value function $V$ as

\begin{eqnarray}\label{eqn:value}
    \min_G \max_D V(D,G) &=& \mathbb{E}_{\mathbf{x}\sim p_{data}}\log D(\mathbf{x})                                                           \nonumber \\
                         &+& \mathbb{E}_{G(\mathbf{z}) \sim p_{model}}\log(1-D(G(\mathbf{z}))) \nonumber \\
\end{eqnarray}

@@

The Nash equilibrium is satisfied by a pair of parameter sets $(\bm{\theta}^{(D)*}, \bm{\theta}^{(G)*})$ for which 
no further changes occur to the objective functions $J^{(D)}$ and $J^{(G)}$ of both networks~\cite{NIPS16}. 
Equation~\ref{eqn:value} means that the zero-sum game for the models returns an adjustment $+V$ for the discriminator 
and $-V$ for the generator. Figure~\ref{fig:gan_game} shows the equilibrium point in the parameter space and the 
returns $V(\bm{\theta}^{(D)}, 0)$ and $V(0, \bm{\theta}^{(G)})$ for the discriminator and generator with optimum 
parameters $\bm{\theta}^{(D)*}$ and $\bm{\theta}^{(G)*}$ respectively in the zero-sum game. The zero-sum game between 
$D$ and $G$ defines an objective function $J^{(D)}$ for the discriminator

\begin{eqnarray}\label{eq:J_D}
    J^{(D)} &=& \mathbb{E}_{\mathbf{x} \sim p_{data}}\log D(\mathbf{x})  \nonumber \\
            &+&\mathbb{E}_{G(\mathbf{z}) \sim p_{model}}\log (1 - D(G(\mathbf{z})))
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Experimental design of GANs} 
The opposing players in the zero-sum game are assumed by two deep convolutional neural network (DCNN) models. These 
networks use convolutional layers to extract features from grid-like data. Deep Convolutional Generative Adversarial 
Networks (DCGANs), first implemented by Radford et al. 2015~\cite{dcgan}, use these layers to create images. The DCGAN 
model has shown excellent results on natural image generation using the CelebA~\cite{celebA} and the 
LSUN-Bedrooms~\cite{LSUN} datasets. The DCNNs in this work are based on the AlexNet models for image 
classification~\cite{dcnn}. Despite most of the architectural features, such as convolutional layers, being used in deep 
learning models prior to this publication the DCGAN architecture improved previous GAN architectures for increased 
performance in image generation tasks. The models in the methods of this work borrow from the ideas of this architectural 
basis.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Applications of an artificial universe}\label{sec:universe_applications}
% WHY IS THIS IMPORTANT, WHAT COULD IT BE FOR, WHAT RESEARCH WILL COME OF IT

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Cosmological simulations}
The Millennium Project simulation~\cite{millsim} and the successive Millennium-XXL~\cite{millxxlsim} and 
Millennium-II~\cite{mill2sim} simulations made foundations for the comparison of observation to theory-based simulation. 
Research using the simulations has provided new insight into high-redshift quiescent galaxies~\cite{mill_quiescent},
dark-matter haloes~\cite{mill_dm_haloes}, weak lensing~\cite{mill_weak_lensing}, galaxy clustering~\cite{mill_gal_clustering} 
and the direct comparison of galaxy surveys with galaxy groups predicted in the simulation volumes\cite{mill_sdss}. 
These applications in the GAN-generated galaxy distributions depend on extending the depth of the method presented in 
this report but they are strong motivations for the use of GANs in simulating cosmological data. The direct comparison 
of survey data and generated data is the more likely once the objective of this work is completed. The approach in this 
report can generate a similar volume to the Millennium simulation in around an hour (GPU time) as opposed to 
three-hundred years (CPU time) for the Millennium simulation.\footnotemark.

\footnotetext{\url{wwmpa.mpa-garching.mpg.de/galform/virgo/millennium}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The process involves the creation of a series of mock data challenges (MDCs). The MDCs are simulations of GW events 
in mock galaxy catalogues. Between the separate MDCs there are varying conditions of GW and EM selection effects 
(GW selection effects correspond to detector sensitivity and EM effects are due to observed flux limitations). The 
effects of clustering and large scale cosmological structure are ignored in the simulation of co-moving volumes of 
uniformly distributed galaxies that the GW events occur in. These mock catalogs show different comparisons to 
certain properties of true galaxy distributions from surveys for locating GW events, such as GLADE~\cite{GLADE}. %For 
% the complete galaxy catalogue in one such MDC a uniform simulated distribution of galaxies is added past the limit 
% of the catalog. 
A larger testing volume could be generated from a routine, such as the one described in this report, with the summary 
statistics that are similar to survey or simulation data with added realistic cosmological structure. The EM 
information from outside the volume specified in the catalogs could not be replaced by the GAN method shown in this 
report because at present it only returns positional data. This could increase the understanding of how current 
techniques for observed GW events perform so that tighter constraints on the source-observer separation could be 
obtained in the future. This would help to accumulate more data for posteriors to further constrain $H_0$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Extraction of training data} % TRAINING DATA MANUFACTURE AND PREPROCESSING
Position vector components for the centres of cube samples in the simulation volume were drawn from a uniform distribution 
along each axis of each redshift-snapshot. If the addition or subtraction of half the side length of the cube to any of 
the components resulted in a point outside of the simulation volume, the point was discarded. The galaxy cube samples were 
obtained by sorting the position vectors of galaxies in the simulation that were within the cube volume. Each point inside 
the cube was recorded in an array. From each simulation volume of $512 \: h^{-1} \,\textup{Mpc}$ ensembles of 64 array 
samples of size $80 \: h^{-1} \,\textup{Mpc}$ were selected. The galaxy positions in each array for every cube were scaled
to the range $[-1,+1]$. 3D-histograms with 32 bins per axis were made on each cube across the same range for each axis. 
The densities in each bin cell were also scaled to this range to fit the output ranges of the GAN models. 
% After testing different sizes, a cube size of $160 \: h^{-1} \,\textup{Mpc}$ was found to balance the selection of unique 
% samples in the simulation volume against featuring the same features in the simulation structure.

@@

where $a$ is a free parameter that depends on the sparsity of the histogram densities. The value of $a=1.25$ was found 
to be optimal for the $160 \:  \: h^{-1} \,\textup{Mpc}$ histograms at all redshifts $z$. This parameter remains fixed 
during training. The value of $a$ controls the median pixel density of the scaled image. Figure \ref{fig:scaling} shows 
the difference between scaled images $s(x)$ and the original images $x$. The inverse transformation $s^{-1}(x)$ can be 
applied to obtain the original image. 
% A logarithmic scaling was used originally, but this did not provide good results with the GAN architecture that used 
% the hyperbolic tangent ($\tanh$) activation because the logarithm of a negative number is undefined. The models using 
% $\tanh$ activations achieved much better results than models without these activations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \subsubsection{2D-GAN implementation} % DESCRIBE SPECIFICS OF THIS IMPLEMENTATION
% \textit{Architecture} --- 
% The networks follow the structuring of the DCGAN models~\cite{dcgan} and are written in Keras~\cite{keras}. A full 
% description of the layers in the 2D-GAN architecture is given in Table~\ref{table:2dgan_archi} of 
% Appendix~\ref{appendix:gan_setups}. The generator transforms a 100-dimensional vector drawn from a Gaussian prior 
% $\mathbf{z}\sim \mathcal{N} (0,1)$ into a $32\times 32$ single-channel image. The latent-space vector is followed by a 
% fully connected layer that is reshaped to a stack of feature maps for the two convolutional layers. Batch-normalisation 
% is only used after the fully connected layer. Through each layer hyperbolic tangent activations are used. See Appendices~\ref{appendix:app_acts} and~\ref{app_BN} for more detail on batch-normalisation and activation functions.

% The discriminator network has two convolutional layers and two fully connected layers. The image samples have features 
% extracted by the first two convolutional layers and the features are flattened for the final fully connected layers. 
% Hyperbolic tangent functions are used throughout the model and a sigmoid activation function (see 
% Appendix~\ref{appendix:app_acts}) is used for the final single neuron layer that outputs the probability for a given 
% sample being real.

% Both networks use the same number of filters for the corresponding layers in either model. Together the discriminator 
% and generator models have $3.8\times 10^7$ trainable parameters. The discriminator and generator objectives of 
% Equations~\ref{eq:J_D} and \ref{eqn:J_G} are minimized using the Adam optimiser~\cite{adam}. Each model uses a learning 
% rate of $2\times 10^{-4}$ and momentum $\beta_1 = 0.5$. The models use a batch size of 64 histograms out of a set of 128 
% training histograms, where the batch size is the number of samples given to the discriminator at each iteration of 
% training. Refer to Table~\ref{table:2dgan_params} of Appendix~\ref{appendix:gan_setups} for a complete listing of the 
% 2D-GAN hyper-parameters used in each model. The hyper-parameters of a model are set outside of training and they shape 
% how the training proceeds.\\

% {\setlength{\parindent}{0cm}
% \textit{Training} --- GANs are notoriously difficult to train. This means that the state of the models can generate 
% realistic samples for a single epoch at a time. The GAN was made to record the parameters of the generator model at 
% these epochs by using a statistical check (see Section~\ref{sec:stats_and_diags}) of the generated histograms against 
% the training data histograms in the training loop. The loss (the objective function value) and accuracy values of the 
% discriminator for each epoch were recorded during training.
% % }

% % An indicator for the performance of a GAN is the classification accuracy of the discriminator model. This metric is 
% the average classification accuracy of the discriminator on two separate batches of real and fake samples. The accuracy 
% metric for the generator is given by the same accuracy subtracted from one. However, in this case the accuracy is the 
% discriminator's accuracy in classifying a batch of fake cubes labelled as real. This is because the generator is trying 
% to make the discriminator classify its products as real. The best results were obtained when the losses of the networks 
% diverged and the accuracies of the networks slowly increased and were at approximately the same value. A training loop 
% of $3.0\times 10^4$ epochs for the $80\:h^{-1} \textup{ Mpc}$ cubes sectioned into $32 \times 32$ pixel projections 
% took approximately 2 hours on an NVIDIA Tesla V100 GPU with 32GB capacity.

@@

\textit{Architecture} --- 
Table~\ref{table:3dgan_archi} of Appendix~\ref{appendix:gan_setups} shows the architectures of the 3D-GAN discriminator 
and generator models. The hyper-parameters used in the 3D-GAN are also given in Table~\ref{table:3dgan_params} of Appendix~\ref{appendix:gan_setups}. Together the discriminator and generator models had around $2.5\times10^8$ trainable 
parameters across one linear layer and three convolutional layers for the generator model and three convolutional layers 
for the discriminator model. \\

{\setlength{\parindent}{0cm}
\textit{Training} --- 
The networks follow the structuring of the DCGAN models~\cite{dcgan} and are written in Tensorflow~\cite{tensorflow} and 
Keras~\cite{keras}. The generator transforms a 256-dimensional vector drawn from a Gaussian prior $\mathbf{z}\sim 
\mathcal{N}(0,1)$ into a $32 \times 32 \times 32$ single-channel image. The latent-space vector is followed by a fully 
connected layerthat is reshaped to a stack of feature maps for the convolutional layers. 
% Through each layer hyperbolic tangent activations are used. See Appendices~\ref{appendix:app_acts} and~\ref{app_BN} 
% for more detail on batch-normalisation and activation functions.
